---
title: "SVFs and GVFs in PSO"
author: "Caden Hewlett"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Slowly-Varying Functions

A function $\ell(x)$ is said to be a Slowly Varying Function (SVF) as $x \rightarrow \infty$ if for any constant $c > 0$
$$
\lim_{x \rightarrow \infty} \Big( \dfrac{\ell(cx)}{\ell(x)}\Big) =1 
$$
The authors (link) use $\ell(x) = ( 10 \ln(x) )^{\alpha}$. We can verify that $\ell$ is an SVF as follows:
$$
\begin{aligned}
\lim_{x \rightarrow \infty} \Big( \dfrac{\ell(cx)}{\ell(x)}\Big) &= \lim_{x \rightarrow \infty} \Big( \dfrac{10^{\alpha} \ln (cx)^{\alpha}}{10^{\alpha} \ln (x)^{\alpha}}\Big) \\
\lim_{x \rightarrow \infty} \Big( \dfrac{\ell(cx)}{\ell(x)}\Big) &= \lim_{x \rightarrow \infty} \Big( \dfrac{\ln(x)^{\alpha}+\ln (c)^{\alpha}}{ \ln (x)^{\alpha}}\Big)
\\
\lim_{x \rightarrow \infty} \Big( \dfrac{\ell(cx)}{\ell(x)}\Big) &\overset{x \text{ dominates}}{=} \lim_{x \rightarrow \infty} \Big( \dfrac{\ln(x)^{\alpha}}{ \ln (x)^{\alpha}}\Big) = \boxed{1}
\end{aligned}
$$

## Regular-Varying Functions (RVF)

A Regular Varying Function (RVF) is one that behaves like a power of $x$, meaning that it grows at a regular rate as $x$ increases. Mathematically, a function $f(x)$ is regularly varying if it satisfies:
$$
\lim_{x \rightarrow \infty} \Big( \dfrac{f(cx)}{f(x)}\Big) = c^{\delta} 
$$
For some constant $\delta$.The authors let $L(x) = x^{\delta} \ell(x)$ as an RVF. 
This makes sense as an RVF because multiplying an SVF by $x^{\delta}$ transforms the slowly varying function into one that grows or shrinks at a regular rate. The control parameter $\delta$ lets you fine-tune this behavior to promote faster or slower convergence depending on the stage of the optimization process.

## In Update Equations

The SVF $\ell(x)$ and RVF $L(x)$ are used in the PSO Update Equations based on $t$. For small $t$, the RVF is used to perturb the $\mathbf{X}$-values as follows:
$$
\mathbf{X}_{t + 1}^{(i)} = \begin{cases} \mathbf{X}_t^{(i)}  + \mathbf{V}_t^{(i)} + \text{RVF}(\mathbf{X}_t^{(i)}), & \text{ if } t \leq \lceil\rho T\rfloor \\
\mathbf{X}_t^{(i)}  + \mathbf{V}_t^{(i)} + \text{SVF}(\mathbf{X}_t^{(i)}), & \text{ otherwise}\end{cases}
$$
Where $t$ is the current iteration, $T$ is the max iteration and $\rho$ is a control parameter dictating when the update perturbations switch from RVF to SVF.


## Implementation:

```{r}
SVF <- function(x, alpha = 0.25){
  return((10 * log(x))^(alpha))
}
x = seq(from = 1, to = 10, length.out = 1000)
plot( SVF(x), type = 'l', ylim = c(0, 7) )

RVF <- function(x, delta = 0.5){
  return( x^(delta)*SVF(x))
}

lines( RVF(x) , col = 'red')
```
